---
title: "EDA: Characterizing the relationship between the religious affiliation and incidences of Covid-19 at U.S. universities."
author: "Daniel de Castro and Laura Appleby"
date: "November 22, 2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, eval=T, include=F}
library(dplyr)
```

For this exploratory data analysis, the first step is to read our data from CSV files into R data frames. The `colleges` data frame stores the NYT data on Covid cases at universities, while the `ipeds` data frame stores the data will most of our predictor variables (university characteristics) taken from IPEDS. We then rename most of the columns in `ipeds` to make them shorter and easier to work with. 

```{r, eval=T}
# Read colleges from csv
colleges <- read.csv("data/colleges.csv")[,-c(1,9)]

# Read ipeds from csv, drop unnecessary columns
ipeds <- read.csv("data/ipeds.csv")
ipeds <- ipeds[-c(3, 8, 22)]

# Rename columns of ipeds 
ipeds <- ipeds %>% rename_at('IC2021.Institutional.control.or.affiliation', 
                             ~'control') %>% 
                    rename_at('HD2021.FIPS.state.code', ~'FIPS.state.code') %>% 
                    rename_at('unitid', ~'ipeds_id') %>% 
                    rename_at('IC2021.Religious.affiliation', 
                              ~'religious.affiliation') %>%
                    rename_at('DRVIC2021.Tuition.and.fees..2020.21', 
                              ~'tuition') %>% 
                    rename_at('DRVEF122021.Total.12.month.unduplicated.headcount', 
                              ~'total.headcount') %>% 
                    rename_at('DRVEF122021.Undergraduate.12.month.unduplicated.headcount',
                    ~'undergrad.headcount') %>% 
                    rename_at('DRVEF122021.Percent.of.12.month.unduplicated.headcount.that.are.American.Indian.or.Alaska.Native', 
                              ~'percent.american.native') %>% 
                    rename_at('DRVEF122021.Percent.of.12.month.unduplicated.headcount.that.are.Asian', 
                              ~'percent.asian') %>% 
                    rename_at('DRVEF122021.Percent.of.12.month.unduplicated.headcount.that.are.Black.or.African.American', 
                              ~'percent.black') %>% 
                    rename_at('DRVEF122021.Percent.of.12.month.unduplicated.headcount.that.are.Hispanic.Latino', ~'percent.hispanic.latino') %>% 
                    rename_at('DRVEF122021.Percent.of.12.month.unduplicated.headcount.that.are.Native.Hawaiian.or.Other.Pacific.Islander', 
                              ~'percent.pacific.islander') %>% 
                    rename_at('DRVEF122021.Percent.of.12.month.unduplicated.headcount.that.are.White', 
                              ~'percent.white') %>% 
                    rename_at('DRVEF122021.Percent.of.12.month.unduplicated.headcount.that.are.two.or.more.races', 
                              ~'percent.two.more.races') %>% 
                    rename_at('DRVEF122021.Percent.of.12.month.unduplicated.headcount.that.are.race.ethnicity.unknown', ~'percent.NA.race') %>% 
                    rename_at('DRVEF122021.Percent.of.12.month.unduplicated.headcount.that.are.Nonresident.Alien',
                              ~'percent.nonres.alien') %>% 
                    rename_at('DRVEF122021.Percent.of.12.month.unduplicated.headcount.that.are.women', ~'percent.women') %>% 
                    rename_at('SFA2021.Average.amount.of.grant.and.scholarship.aid.awarded..2020.21', ~'avg.grant.money') %>% 
                    rename_at('DRVGR2021.Graduation.rate..total.cohort', ~'grad.rate') %>% 
                    rename_at('SFA2021.Percent.of.full.time.first.time.undergraduates.awarded.any.financial.aid', 
                              ~'percent.fin.aid') %>% 
                    rename_at('SFA2021.Percent.of.full.time.first.time.undergraduates.awarded.student.loans', 
                              ~'percent.student.loan') %>% 
                    rename_at('IC2021.Occupational', ~'occupational.degree') %>% 
                    rename_at('IC2021.Academic', ~'academic.degree') %>% 
                    rename_at('IC2021.Adult.basic.remedial.or.high.school.equivalent', 
                              ~'hs.equivalent.degree') %>% 
                    rename_at('IC2021.Percent.of.undergraduates..who.are.formally.registered.as.students.with.disabilities..when.percentage.is.more.than.3.percent', 
                              ~'percent.disability') %>% 
                    rename_at('IC2021.NCAA.NAIA.conference.number.football', 
                              ~'NCAA.football') %>% 
                    rename_at('IC2021.Institution.provide.on.campus.housing', 
                              ~'on.campus.housing') %>% 
                    rename_at('IC2021.Total.dormitory.capacity', 
                              ~'dorm.capacity') %>% 
                    rename_at('IC2021.Typical.room.charge.for.academic.year', 
                              ~'dorm.room.price')
```



```{r, eval=T}
# Merge data frames, and remove duplicated "state" column 
md <- merge(ipeds, colleges, by="ipeds_id")[,-32]

# Remove institutions with no ipeds data (a lot of NAs)
md <- md[!(md$control == ""),]

# Create column 'religious' 
md$religious <- "Yes"
md[md$religious.affiliation == "Not applicable", ]$religious <- "No"

# Create column 'catholic'
md$catholic <- "Non-Catholic"
md[md$religious.affiliation == "Not applicable",]$catholic <- "No"
md[md$religious.affiliation == "Roman Catholic",]$catholic <- "Yes"

# Make 'control' column simpler
md[md$control == "Private not-for-profit (no religious affiliation)",]$control <- "Private not-for-profit"
md[md$control == "Private not-for-profit (religious affiliation)",]$control <- "Private not-for-profit"

# Write to csv 
write.csv(md,'merged_cases.csv')
```

## Daniel's work 

```{r, eval=T}
mask.mandates <- read.csv("data/U.S._State_and_Territorial_Public_Mask_Mandates_From_April_8__2020_through_August_15__2021_by_State_by_Day.csv")
factor.variables <- c("Face_Masks_Required_in_Public", "State_Tribe_Territory", "order_code")
mask.mandates[,factor.variables] <- lapply(mask.mandates[,factor.variables], as.factor)
mask.mandates$date <- as.Date(mask.mandates$date, format="%m/%d/%Y")
```

```{r, eval=T}
mandates.by.state <- data.frame(state=unique(mask.mandates$State_Tribe_Territory))
mandates.by.state$number_of_days_mandated <- 0
for (i in 1:length(unique(mandates.by.state$state))) {
  dummy <- mask.mandates$Face_Masks_Required_in_Public[mask.mandates$State_Tribe_Territory == mandates.by.state$state[i]
                                                       & mask.mandates$date >= as.Date("2020/7/1")
                                                       & mask.mandates$date <= as.Date("2021/5/26")]
  dummy[is.na(dummy)] <- "No"
  mandates.by.state$number_of_days_mandated[mandates.by.state$state == mandates.by.state$state[i]] = sum(dummy == "Yes")
}
mandates.by.state$full_state_name <- c("Alaska", "Alabama", "Arkansas", 
                                       "American Samoa", "Arizona", 
                                       "California", "Colorado", 
                                       "District of Columbia", "Connecticut", 
                                       "Florida", "Delaware", "Georgia", "Guam", 
                                       "Iowa", "Hawaii", "Idaho", "Illinois", 
                                       "Indiana", "Kansas", "Kentucky", 
                                       "Louisiana", "Massachusetts", 
                                       "Minnesota", "Maryland", "Maine", 
                                       "Michigan", "Missouri", 
                                       "Northern Mariana Islands", 
                                       "Mississippi", "Montana", 
                                       "North Carolina", "North Dakota", 
                                       "Nebraska", "New Hampshire", "Nevada", 
                                       "New Jersey", "New Mexico", "Ohio", 
                                       "New York", "Oklahoma", "Oregon", 
                                       "South Carolina", "Rhode Island", 
                                       "Pennsylvania", "Puerto Rico", 
                                       "South Dakota", "Tennessee", "Texas", 
                                       "Utah", "Virginia", "Virgin Islands", 
                                       "Vermont", "Washington", "Wisconsin", 
                                       "West Virginia", "Wyoming")
mandates.by.state <- mandates.by.state[,c(3,2)]


full.cases <- merge(md, mandates.by.state, by.x="FIPS.state.code", 
      by.y="full_state_name")
full.cases$total.cases <- full.cases$cases + full.cases$cases_2021

# Write to csv
write.csv(full.cases, file="data/full_cases.csv")
```

## Laura's work 

```{r, eval=T}
# TODO: CONSIDER CHANGING THIS EPSILON VALUE TO GIVE BETTER PLOTS 
epsilon <- 0.001

hist(full.cases$cases)
hist(log(full.cases$cases + epsilon))

hist(full.cases$cases_2021)
hist(log(full.cases$cases_2021 + epsilon))

hist(full.cases$total.cases)
hist(log(full.cases$total.cases + epsilon))

# this shows that religious institutions are not as large as the largest non-religious
# this is because of large public non-religious schools
# once accounting for size, median case count higher for religious.
boxplot((total.cases  / total.headcount) ~ religious, data=full.cases)

# Convert necessary columns to factors 
factor.variables <- c("religious", "FIPS.state.code", "control", "occupational.degree", "academic.degree", "hs.equivalent.degree", "NCAA.football", "on.campus.housing", "catholic")
full.cases[,factor.variables] <- lapply(full.cases[,factor.variables], as.factor)
```

```{r, eval=T}
summary(lm1 <- lm(total.cases ~ religious, full.cases))
summary(lm2 <- lm(log(total.cases + epsilon) ~ religious, full.cases))

summary(lm3 <- lm(log(total.cases + epsilon) ~ religious + FIPS.state.code + control + percent.fin.aid + tuition + number_of_days_mandated + total.headcount, full.cases))

lm4.predictors <- c("religious", "FIPS.state.code", "control", "percent.fin.aid", "tuition", "number_of_days_mandated", "total.headcount")
summary(lm4 <- lm(log(total.cases + epsilon) ~ (religious + FIPS.state.code + control + percent.fin.aid + tuition + number_of_days_mandated)*total.headcount, data=full.cases[complete.cases(full.cases[,lm4.predictors]),]))
# NA predictors b/c of singularities: FIPS.state.code and number_of_days_mandated are highly collinear,
# as the latter depends on the former 
```










